---
title: "Portfolio Analysis"
author: Sámuel Császár
abstract: | 
  This report uses Monte Carlo simulation to compare the 10 day Value at Risk  and Conditional Value at Risk of a diversified equity portfolio against a concentrated technology portfolio, both equally weighted across 11 large cap stocks. The diversified portfolio's simulated risk was consistently about half that of the concentrated portfolio, a gap that is consistent with a lower average correlation among its holdings. Results were validated through historical backtesting. Both portolios' return distributions were found to be leptokurtic, which is a source of tail risk that is not captured by the model's assumption of normality.
format:
  html:
    theme: journal
    toc: true
code-fold: true
echo: true
output: false
---

## Executive Summary

This analysis compares the risk profiles of two equally weighted equity portfolios (containing 11 stocks each) over a five year period from 2021 to 2026. The first portfolio is diversified, containing stocks from a variety of sectors (technology, energy, healthcare, financials, consumer staples, and industrials). The second portfolio is concentrated containing 11 large cap technology stocks. the diversified portfolio exhibited a significantly lower average pairwise correlation between holdings (0.25 compared to 0.45). Moreover, at a 10 day horizon, both its Value at Risk (VaR) and Conditional Value at Risk (CVaR) were roughly half that of the concentrated portfolio at both 95% and 99% confidence intervals (e.g. 95% VaR of 4.37% compared to 8.65%).

```{r}
#| code-summary: "Packages Used"

library(tidyverse)
library(ggplot2)
library(tidyquant)
library(patchwork)
```

```{r}
#| code-summary: "Constructing the Portfolios"

tickers_A <- c("AAPL", "MSFT", "XOM", "CVX", "JNJ", "PFE", "JPM", "BAC", "PG", "KO", "CAT")
tickers_B <- c("AAPL", "MSFT", "NVDA", "GOOGL", "META", "AMZN", "AVGO", "CRM", "ADBE", "ORCL", "AMD")
start_date <- "2021-01-01"
end_date <- "2026-09-07"
portfolio_A <- tq_get(tickers_A,
       from = start_date, 
       to = end_date,
       get = "stock.prices")
portfolio_B <- tq_get(tickers_B,
       from = start_date, 
       to = end_date,
       get = "stock.prices")
portfolio_A |> 
  group_by(symbol) |> 
  summarise(n = n(), n_na = sum(is.na(adjusted)))
portfolio_B |> 
  group_by(symbol) |> 
  summarise(n = n(), n_na = sum(is.na(adjusted)))

```

```{r}
portfolio_A <- portfolio_A |> 
  group_by(symbol) |> 
  arrange(date) |> 
  mutate(daily_return = log(adjusted / lag(adjusted))) |>
  ungroup()

portfolio_B <- portfolio_B |> 
  group_by(symbol) |> 
  arrange(date) |> 
  mutate(daily_return = log(adjusted / lag(adjusted))) |>
  ungroup()

portfolio_A |> 
  group_by(symbol) |> 
  summarise(mean_return=mean(daily_return, na.rm = T),
            sd_return=sd(daily_return, na.rm = T),
            .groups = "drop")

portfolio_B |> 
  group_by(symbol) |> 
  summarise(mean_return=mean(daily_return, na.rm = T),
            sd_return=sd(daily_return, na.rm = T),
            .groups = "drop")
```

```{r}
returns_wide_A <- portfolio_A |> 
  filter(!is.na(daily_return)) |> 
  select(date, symbol, daily_return) |> 
  pivot_wider(names_from = symbol, values_from = daily_return)

returns_wide_B <- portfolio_B |> 
  filter(!is.na(daily_return)) |> 
  select(date, symbol, daily_return) |> 
  pivot_wider(names_from = symbol, values_from = daily_return)
```

```{r}

cor_matrix_A <- cor(returns_wide_A |> 
                      select(-date), 
                    use = "pairwise.complete.obs")
cor_matrix_B <- cor(returns_wide_B |> 
                      select(-date), 
                    use = "pairwise.complete.obs")
cor_matrix_A
cor_matrix_B
avg_cor <- function(mat){
  mean(mat[upper.tri(mat)])
}
avg_cor(cor_matrix_A)
avg_cor(cor_matrix_B)
min(cor_matrix_A[upper.tri(cor_matrix_A)])
min(cor_matrix_B[upper.tri(cor_matrix_B)])
max(cor_matrix_A[upper.tri(cor_matrix_A)])
max(cor_matrix_B[upper.tri(cor_matrix_B)])
which(cor_matrix_A == max(cor_matrix_A[upper.tri(cor_matrix_A)]), 
      arr.ind = TRUE)
which(cor_matrix_B == max(cor_matrix_B[upper.tri(cor_matrix_B)]), 
      arr.ind = TRUE)

plot_corr_heatmap <- function(cor_matrix, title){
  cor_df <- as.data.frame(cor_matrix) |> 
    mutate(symbol1 = rownames(cor_matrix)) |> 
    pivot_longer(-symbol1, names_to = "symbol2",
                 values_to = "correlation")
  ggplot(cor_df, aes(x = symbol1, y = symbol2, 
                     fill = correlation))+
    geom_tile()+
    scale_fill_gradient2(low = "blue", mid = "white",
                         high = "red", midpoint = 0,
                         limits = c(-1,1))+
    labs(title = title, x = NULL, y = NULL)+
    theme_minimal()+
    theme(axis.text.x = element_text(angle = 45, 
                                     hjust = 1))
  
}

```

```{r}
#| fig-height: 6
#| fig-width: 12
#| output: true
plot_corr_heatmap(cor_matrix_A, 
                  "Diversified Portfolio - Correlation")+
plot_corr_heatmap(cor_matrix_B, "Concentrated Tech Portfolio - Correlation")
```

```{r}
library(MASS, exclude = "select")
set.seed(123)
simulate_portfolio_var <- function(returns_wide, n_sims = 10000,
                                   horizon_days, 
                                   confidence_levels = c(0.95, 0.99)){
  returns_matrix <- returns_wide |> 
    dplyr::select(-date) |> 
    as.matrix()
  n_assets <- ncol(returns_matrix)
  weights <- rep(1/n_assets, n_assets)
  mu <- colMeans(returns_matrix, na.rm = T)
  sigma <- cov(returns_matrix, use = "pairwise.complete.obs")
  simulated_returns <- mvrnorm(n = n_sims, mu = mu, Sigma = sigma)
  simulated_portfolio_returns <- simulated_returns %*% weights
  simulated_portfolio_returns <- simulated_portfolio_returns * sqrt(horizon_days)
  results <- lapply(confidence_levels, function(cl){
    var_threshold <- quantile(simulated_portfolio_returns, probs = 1-cl)
    cvar <- mean(
      simulated_portfolio_returns[simulated_portfolio_returns<=var_threshold])
    tibble(confidence = cl, VaR = -var_threshold, CVaR = -cvar)
  }) |> 
    bind_rows()
  list(
    simulated_returns = simulated_portfolio_returns,
    summary = results
  )
}
var_results_A <- simulate_portfolio_var(returns_wide_A, 
                                        horizon_days = 10)
var_results_B <- simulate_portfolio_var(returns_wide_B, 
                                        horizon_days = 10)
var_results_A$summary
var_results_B$summary

```

```{r}
historical_portfolio_returns_A <- returns_wide_A |> 
  select(-date) |> 
  as.matrix() %*% rep(1/11,11)

realized_10d_A <- zoo::rollapply(
  historical_portfolio_returns_A, 
                               width = 10,
                               FUN = sum,
                               align = "right")
var95_A <- var_results_A$summary$VaR[
  var_results_A$summary$confidence == 0.95]
breaches_A <- sum(realized_10d_A < -var95_A)
breaches_rate_A <-breaches_A/length(realized_10d_A)
breaches_rate_A

var_99_A <- var_results_A$summary$VaR[
  var_results_A$summary$confidence == 0.99]

breaches_99_A <- sum(realized_10d_A < -var_99_A)
breach_rate_99_A <- breaches_99_A / length(realized_10d_A)

breach_rate_99_A

historical_portfolio_returns_B <- returns_wide_B |> 
  select(-date) |> 
  as.matrix() %*% rep(1/11,11)

realized_10d_B <- zoo::rollapply(
  historical_portfolio_returns_B, 
                               width = 10,
                               FUN = sum,
                               align = "right")
var95_B <- var_results_B$summary$VaR[
  var_results_B$summary$confidence == 0.95]
breaches_B <- sum(realized_10d_B < -var95_B)
breaches_rate_B <-breaches_B/length(realized_10d_B)
breaches_rate_B

var_99_B <- var_results_B$summary$VaR[
  var_results_B$summary$confidence == 0.99]

breaches_99_B <- sum(realized_10d_B < -var_99_B)
breach_rate_99_B <- breaches_99_B / length(realized_10d_B)

breach_rate_99_B

library(e1071)
kurtosis(historical_portfolio_returns_A)
kurtosis(historical_portfolio_returns_B)
```

```{r}
#| output: true
sim_df <- bind_rows(
  tibble(portfolio = "Diversified", return = as.numeric(
    var_results_A$simulated_returns)),
  tibble(portfolio = "Concentrated Tech", return = as.numeric(
    var_results_B$simulated_returns))
)

var_lines <- bind_rows(
  var_results_A$summary |> 
    mutate(portfolio = "Diversified"),
  var_results_B$summary |> 
    mutate(portfolio = "Concentrated Tech")
)

ggplot(sim_df, aes(x = return, fill = portfolio))+
  geom_density(alpha = 0.5)+
  geom_vline(data = var_lines |> 
               filter(confidence == 0.95), aes(xintercept = -VaR, 
                                                color = portfolio),
             linetype = "dashed", linewidth = 0.8)+
  scale_fill_manual(values = c("Diversified" = "steelblue",
                               "Concentrated Tech" = "firebrick"))+
  scale_color_manual(values = c("Diversified" = "steelblue",
                                "Concentrated Tech" = "firebrick"))+
  labs(title = "Simulated 10 day Portfolio Returns", 
       subtitle = "Dashed lines mark 95% VaR threshold",
       x = "10 Day Return", y = "Density", fill = "Portfolio")+
  guides(color = "none")+
  theme_minimal()

```

```{r}

```

